An Efficient Feedback Calibration Algorithm for Direct Imaging Radio Telescopes

### 2016-06-20 cal_demo/ms.tex
Fine tuning notation, language, and consistency. Ready to write up response to referee

% Abstract of the paper
\begin{abstract}
We present the E-field Parallel Imaging Calibration (EPICal) algorithm, which addresses the 
need for a real-time calibration method for direct imaging radio astronomy correlators. Direct 
imaging involves a spatial fast Fourier transform of antenna voltages, alleviating the harsh 
$\mathcal{O}(\Nant^2)$ computational scaling to a more gentle $\mathcal{O}(\Nant \log_2 
\Nant)$, which can save orders of magnitude in computation cost for next generation arrays 
consisting of hundreds to thousands of antennas. However, because signals are mixed in the 
correlator, gain correction must be applied on the front end. We develop the EPICal algorithm 
to form gain solutions in real time without ever forming visibilities. This method scales as the 
number of antennas, and produces results comparable to those from visibilities. Through 
simulations and application to Long Wavelength Array data we show this algorithm is a 
promising solution for next generation instruments.
\end{abstract}

@@

\section{Introduction}
In order to satisfy the survey speeds required for precision cosmology as well as searches for 
fast radio transients, radio astronomy is undergoing a paradigm shift toward interferometers 
consisting of hundreds to thousands of small, widefield antennas. Many arrays with this design 
are already built or under construction including the Hydrogen Epoch of Reionization Array
\footnote{http://reionization.org} (HERA), the Murchison Widefield Array (MWA; 
\citealt{tin13,bow13}), the Donald C. Backer Precision Array for Probing the Epoch of 
Reionization (PAPER; \citealt{par10}), the LOw Frequency ARray (LOFAR; \citealt{van13}), the 
Canadian Hydrogen Intensity Mapping Experiment (CHIME,\citealt{ban14}), the Long 
Wavelength Array (LWA, \citealt{ell13}), and the low frequency Square Kilometer Array (SKA1-
Low \citealt{mel13}).

Traditional radio correlators cross-multiply the voltage signals from all pairs of antennas, and 
the computation scales as the number of antennas squared, $\mathcal{O}(\Nant^2)$ 
\citep{bun04}. As the number of elements in future arrays grows, the computational cost will 
become prohibitively expensive, and exploring efficient correlator schemes is essential to 
enable next generation instruments \citep{lon00}. Meanwhile, radio transient monitoring 
requires access to high time and frequency resolution data to identify and characterize events 
such as fast radio bursts (FRBs, \citealt{lor07}), or to follow up gravitational wave candidates 
with radio observations \citep{abb16a,abb16b}. FRBs are highly unexplored at low frequencies 
(< 1 GHz), but are expected to occur on timescales $\Delta t \sim$ 1--10~ms \citep{tho13}. 
Recording the full visibility matrix for $\Nant \gtrsim 10^3$ arrays at this timescale leads to 
extremely high data write rates.

Direct imaging correlators are a new variety of radio correlator which aim to alleviate both the 
computational strain of forming $\Nant^2$ correlations and the high data throughput associated 
with short timescale science. This is done by performing a spatial fast Fourier transform (FFT) 
to image the antenna voltages, then squaring and averaging in time. This process scales as 
$\mathcal{O}(\Ng \log_2 \Ng)$, where \Ng~is the number of grid points in the FFT \citep{mor11,
 teg09, teg10}. For certain classes of telescopes, significantly those envisioned for next 
 generation cosmology experiments, this scaling is a large improvement over the $\Nant^2$ 
 scaling of traditional methods. Furthermore, because images are generated online, the native 
 output bandwidth will be lowered in cases where $\Ng < \Nant^2$, and has the potential to be 
 lowered even further with online transient processing.

@@

Here we present the E-field Parallel Imaging Calibration (EPICal) algorithm -- a novel solution 
to the calibration problem, which can be integrated into direct imaging correlators and scales 
only as the number of antennas, $\mathcal{O}(\Nant)$. This method uses a correlation of the 
uncalibrated antenna signal stream with an output image pixel from the backend of the 
correlator to solve for the complex gains of the antennas. Because the calibration must be 
applied before gridding and imaging, our solution requires an iterative approach where the data 
from one time series is used to update the gains which are applied to the following time series. 
An example implementation of the algorithm is available with the EPIC software package
\footnote{http://github.com/nithyanandan/EPIC}.

We review the MOFF algorithm and derive the calibration algorithm in \S \ref{sec:math}. We 
then demonstrate the algorithm in simulations in \S \ref{sec:sim}, and apply to a sample LWA 
data set in \S \ref{sec:data}. Then we discuss errors and noise properties of the resulting gain 
solutions in \S \ref{sec:noise}. Finally we conclude and discuss potential extensions to the 
algorithm in \S \ref{sec:discussion}.

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{fig1.pdf}
\caption{Data flow for the MOFF algorithm, reproduced from \citealt{thy15c}. The time ordered 
electric field measurements are collected by the antenna elements then frequency transformed in the 
F-engine. The calibration is applied, then the electric field measurements are gridded using the 
antenna illumination patterns. The imager performs a spatial 2D FFT on the gridded data before 
squaring and averaging. The accumulated images are then written to disk.}
\label{fig:moff_flow}
\end{center}
\end{figure}

@@

Next we turn to the measurement performed by an antenna with far-field radiation pattern 
$\mathcal{W}_a(\hat{\mathbf{s}})$, where the subscript $a$ denotes a specific antenna. We 
adopt the common convention of normalizing the radiation pattern to have response of 1 at the 
center of the main lobe \citep[e.g.][]{nap99}. Consequently, the translation between the terminal 
voltages to electric field measurements has been absorbed in our definition of the radiation 
pattern. The ideal antenna electric field measurement can be expressed as 
\begin{equation}
E^I_a(f,t) = \int \mathcal{W}_a(\hat{\mathbf{s}})\mathcal{E}(\hat{\mathbf{s}},f,t) e^{-2\pi i \mathbf{r} \cdot \hat{\mathbf{s}}}\, d\Omega. 
\end{equation}
Equivalently we can express the effect of the antenna radiation pattern in the plane of the 
aperture.
\begin{equation}
E^I_a(f,t) = \int_{\mathrm{aperture}}W_a(\mathbf{r}-\mathbf{r}_a)E_a(\mathbf{r},f,t)\,d^2\mathbf{r}
\end{equation}
Here we used the Fourier relationship between the antenna's far-field radiation pattern and the 
aperture illumination pattern (sometimes referred to as the aperture distribution), $W_a(\mathbf{r}-\mathbf{r}_a)=\int  \mathcal{W}_a(\hat{\mathbf{s}}) \, d
\Omega$. Although any antenna extends over a region of space, we include a nominal antenna 
position, $\mathbf{r}_a$, for multiple reasons. In the case where all antennas of an array are 
identical, a global aperture distribution can be adopted and the translation of the antennas 
distinguishes them. However, even if all antennas have unique aperture distributions, it can be 
convenient to retain the relative positions of antennas in the analysis. This allows for easy 
comparison with visibility-based analyses where the baseline vectors are used to describe 
sampling of the $uv$ plane. From a practical point of view, measurements of the antenna 
radiation pattern may not be performed in-situ, so it is convenient to describe the antenna 
aperture distribution relative to a point fixed within the aperture, along with a translational 
position.

@@

The next step is to grid the calibrated electric field measurements to a finite set of regularly spaced 
positions on the 2D plane of the ground, $\mathbf{r}_j$.
The extent of the grid is chosen to encompass the footprint of the antenna array.
MOFF uses the antenna aperture distributions as the gridding 
kernel according to the software holography/A-transpose technique \citep{mor09,bha08}. 
Mathematically we can express the gridding operation as
\begin{equation}\label{eq:gridding}
E'_j=\sum_a W_a(\mathbf{r}_j-\mathbf{r}_a) E'_a(t)
\end{equation}
Formally each pixel is calculated by summing over all antennas and their illumination patterns.
However the illumination patterns are typically compact and non-overlapping, so most pixels
will only contain contribution from at most one antenna, allowing equation~\ref{eq:gridding}
to be calculated through a sparse matrix multiplication.
This gridding 
achieves an optimal map \citep{teg97b} and places the data on a regular grid. Unlike 
traditional correlators which grid visibilities, the MOFF grids the electric field measurements directly. This 
operation must be performed for every time sample, before any averaging of the data.

@@

We conclude this section by connecting our calibration expression to that found in a visibility 
framework. In the limit of a single bright calibrating source at center, we can greatly 
simplify equations~\ref{eq:Cna} and~\ref{eq:cal_solution}. We will assume the radiation patterns are 
normalized such that $\mathcal{W}(0)=1$. We can further drop the exponential phase terms 
because $\spix$ is perpendicular to our planar antenna array. We then absorb the true gains into the true visibilities in 
equation~\ref{eq:Cna} to express as a sum of measured, uncalibrated, visibilities.
\begin{equation}
\mathcal{C}^{(n)}_{a,0} \rightarrow \frac{1}{\Nant}\sum_b \frac{1}{g^{*(n)}_b} \V_{ab}
\end{equation}

We next plug this expression into equation~\ref{eq:cal_solution} to find our simplified 
calibration solution for a single bright point source. Because our sky is a single bright point 
source, the model visibilities are simply the flux of the source, $S_{\mathrm{src}}$.
\begin{equation}
g^{(n+1)}_a \rightarrow \frac{\sum_b  V_{ab}/g^{*(n)}_b}{\Nant S_{\mathrm{src}}}
\end{equation}
This is simply a gain-weighted sum of the measured visibilities over the flux of the source, 
which is indeed the limiting result from a visibility approach, for example seen in \citealt{mit08}. 
The ability to recover the equivalent expression despite not actually forming the visibilities is a 
result of the fact that only sums over visibilities come into the visibility-based solution, as was described in 
\citealt{mor11}. We have confirmed the limiting case equivalence here, and will explore the 
more general case in more detail in \S~\ref{sec:noise}.

@@

The calibration loop continues by updating the gain estimates every \caliter~time samples. The 
phases of our gain estimates are shown in figure~\ref{fig:sim_phase} for 20 such iterations. 
The phase error plotted is the phase relative to the true gain for each antenna (various colored 
lines). One antenna was used as a reference to fix the absolute phase, so has zero phase 
error. The other 50 antennas are shown to have error spanning $2\pi$ initially, and after about 
10 iterations lock into a solution, settling down to noise levels around iteration 15 (0.15~s). We 
stop the simulation when the updated gains trace the thermal noise of the simulated sources, 
which can be seen by the coherence of the 50 antenna gains after iteration 15.

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{fig3.pdf}
\caption{Phase error of gain estimates as a function of iteration for simulated calibration. The 
gains were initialized with random phases, but the calibration loop was able to recover the 
correct phases after about 15 iterations. Each line represents an antenna in our 51 MWA 
antenna sample.
}
\label{fig:sim_phase}
\end{center}
\end{figure}

The estimated gain amplitudes for the simulations are shown in figure~\ref{fig:sim_amp}. The 
quantity plotted is the magnitude of the estimated gains over the true gains, $\left|g^{(n)}_a/
g^T_a\right|$, which places all antennas on the same scale. We can see the amplitudes 
converge toward their true values around the same time as the phases (iteration $\sim$15). At 
the beginning of calibration we can see the importance of the damping factor. At $n=0$, a couple of 
gains are shown to have abnormally high amplitude estimates, notably one about 3.3 times its 
true value (red line). These unbalanced high estimates caused the entire set of gains to be 
under estimated at $n=1$, even with a damping factor of 0.35. By $n=5$ the unbalanced 
amplitudes have been damped out and the calibration continues. Without the damping factor, 
the oscillation seen in the first couple iterations would have been significantly larger and taken 
much longer to fade out.

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{fig4.pdf}
\caption{Gain estimate amplitudes as a function of iteration for simulated calibration. Again, 
each line represents an antenna in the 51 MWA antenna sample. The gain estimates were 
initialized randomly, while the true values were unity. After about 15 iterations we see the 
calibration loop has settled around the correct values, with only noise remaining.}
\label{fig:sim_amp}
\end{center}
\end{figure}

After the gains have converged, we see both the phases and amplitudes to continue to 
fluctuate coherently. This is due to the stochastic fluctuations of the simulated sources 
themselves, and is limited by the sky noise. For this simulation we restricted each calibration 
iteration to only 10~ms integration time. In principle a calibration implementation could use 
short integration times to allow the gains to converge, then increase the integration time to 
reduce this noise.

Images created at the beginning of calibration and at the end are shown in 
figure~\ref{fig:sim_images}. Each image is obtained over 10~ms integration, corresponding to 
all snapshot images created with a given set of gain estimates. The left panel shows the image 
produced with our initialized unity gains. Because the phases are completely random, the 
image is essentially noise with the primary beam evident. After 20 iterations, the image is far 
more clear, shown in the middle panel. 
The ten sources are visible, along with rumble throughout the image. By comparing with
a perfectly calibrated image (right panel), we see that the rumble and the shapes of the
point sources are matched between the two images. We therefore conclude that the 
artifacts visible are dominated by the point spread function, rather than calibration errors.

\begin{figure*}
\begin{center}
\includegraphics[width=.3\linewidth]{fig5a.pdf}
\includegraphics[width=.3\linewidth]{fig5b.pdf}
\includegraphics[width=.3\linewidth]{fig5c.pdf}
\caption{Images formed during simulated calibration. \emph{Left:} An image generated from a 
single frequency channel and 10 ms integration after the gain estimates are randomized. As 
expected with random phases, the image is completely noisy, with the shape of the primary 
beam evident. \emph{Middle:} An image formed after calibration, again with a single frequency 
channel and 10 ms integration. Now the simulated point sources are visible in the image. 
\emph{Right:} An image generated from the same simulated data as the middle image, but 
with perfect calibration correction. Many of the features seen in the middle image are also
seen here, confirming that they are point spread function artifacts. The ten simulated (and modeled) point sources are highlighted with red circles.
\label{fig:sim_images}
}
\end{center}
\end{figure*}

\section{Noise trends}\label{sec:noise}
Now next study the effect of noise in our system, and the consequences 
of an incomplete sky model. We run a suite of simulations varying the receiver noise and 
integration times, while forming antenna-pixel correlations, $\Cna$, for EPICal gain solutions, 
and simultaneously forming visibilities from the same electric field measurement streams to find visibility-based 
gain solutions for comparison.

@@

To simulate receiver noise, we add a gaussian distributed complex random number to each 
antenna electric field measurement at each time sample according to 
equation~\ref{eq:apply_gain}. The level of the receiver noise, $\sigma_r = \left<\left|
n_a(f,t)\right|^2\right>$, is varied in different simulation runs. We include two limiting 
cases, where the receiver noise power is subdominant to the sky power ($\sigma_r = 
10.0$~Jy), and where the receiver dominates the noise ($\sigma_r=100.0$~Jy).

@@

We simultaneously form simulated visibilities, $V_{ab}$, by correlating all pairs of antenna electric field 
measurements (including the receiver noise). The electric field measurements are correlated for a duration equal to 
the effective integration time of the EPICal loop for comparison. We then find the visibility-
based gain estimates by minimizing
\begin{equation}\label{eq:vis_cal}
\chi^2 = \sum_a\sum_{b\ne a} \left|V_{ab}-g_a g_b^* V^M_{ab}\right|^2,
\end{equation}
using both versions of the model visibilities described above.

@@

We made several choices while studying the behavior of the LWA data to improve our 
calibration. Through our previous imaging work, we noted that the flux scale of uncalibrated 
images was consistent with average gain amplitudes of 0.25. To allow the calibration to 
converge quickly we initialized our gain estimates at this level. 
We also found a boost in signal to noise is achieved easily by assuming the gains are
constant across a range of frequencies, and averaging solutions across channels.
Here we average solutions across 300 channels, or about 58.6 kHz. 
With a fractional bandwidth $B/f_0 = 7.9 \times 10^{-4} \ll 1$, we assume a smooth bandpass 
across the band. 
A damping factor of $\gamma = 0.7$ was adopted.
The large damping factor, relative to what was used in simulations, was necessary due to
the larger noise levels in the LWA data compared to the ideal simulations.

@@

Figure~\ref{fig:data_phase} shows the phase of our gain estimates over 30 calibration 
iterations, again with each colored line representing a different antenna. Given the quality of 
uncalibrated image demonstrated in \cite{thy15c}, it is a bit surprising to see the phase 
variation in our solutions. However, the phases are relatively flat after about 15 iterations 
(modulo noise), and exhibit a central ``trunk" where the majority of phases are congregated. 
This behavior is suggestive that while the uncalibrated voltages were able to produce a viable 
image, the minor changes from our solutions will focus the image and improve the quality. The 
actual location of the ``trunk" (slightly negative) is simply determined by the reference antenna 
chosen to have identically zero phase, but happens to be slightly more positive than the bulk of 
antennas. For plotting clarity, we unwrapped phases resulting in phases that appear to exceed 
$\pm \pi$.

@@

To estimate our fractional gain noise we calculate the standard deviation of the gain
estimates after iteration 15 for each antenna separately. We then divide these standard
deviations by the respective mean gain amplitudes. This resulted in a distribution of fractional
gain noise centered at 0.15 and a width of 0.04 ($\approx 0.15/\sqrt{15}$). This noise level
is consistent with what we expected from the simulations.
There were four outliers with noise levels significantly outside the distribution 
($\sigma_g \approx 0.29$). Upon closer inspection the gain amplitudes of these outliers had
converged later in the calibration loop, after iteration 15.

@@

\textbf{Improving sky model.} In this work we used a perfect sky model (for simulation), or a 
very simple sky model (for LWA data). In principle the images produced by the EPIC correlator 
can be used to improve the sky model used in calibration -- similar to a major loop in self-
calibration. This can be especially useful for compact, widefield arrays which require a model of 
both compact and diffuse sources over a large patch of sky. In addition to requiring the sky 
model, the calibration also requires this model to be attenuated by the primary beam, which 
can be difficult to measure at the precision necessary \citep[e.g.][]{neb15,vir14,thy15b}. 
However, the direct imaging correlator provides exactly the model necessary in real time, and 
can be iterated over to improve the images and gain solutions.

>>>>

% Abstract of the paper
\begin{abstract}
We present the E-field Parallel Imaging Calibration (EPICal) algorithm, which addresses the 
need for a real-time calibration method for direct imaging radio astronomy correlators. Direct 
imaging involves a spatial fast Fourier transform of antenna signals, alleviating the harsh 
$\mathcal{O}(\Nant^2)$ computational scaling to a more gentle $\mathcal{O}(\Nant \log_2 
\Nant)$, which can save orders of magnitude in computation cost for next generation arrays 
consisting of hundreds to thousands of antennas. However, because signals are mixed in the 
correlator, gain correction must be applied on the front end. We develop the EPICal algorithm 
to form gain solutions in real time without ever forming visibilities. This method scales as the 
number of antennas, and produces results comparable to those from visibilities. Through 
simulations and application to Long Wavelength Array data we show this algorithm is a 
promising solution for next generation instruments.
\end{abstract}

@@

\section{Introduction}
In order to satisfy the survey speeds required for precision cosmology as well as searches for 
fast radio transients, radio astronomy is undergoing a paradigm shift toward interferometers 
consisting of hundreds to thousands of small, widefield antennas. Many arrays with this design 
are already built or under construction including the Hydrogen Epoch of Reionization 
Array\footnote{http://reionization.org} (HERA), the Murchison Widefield Array (MWA; 
\citealt{tin13,bow13}), the Donald C. Backer Precision Array for Probing the Epoch of 
Reionization (PAPER; \citealt{par10}), the LOw Frequency ARray (LOFAR; \citealt{van13}), the 
Canadian Hydrogen Intensity Mapping Experiment (CHIME,\citealt{ban14}), the Long 
Wavelength Array (LWA, \citealt{ell13}), and the low frequency Square Kilometer Array (SKA1-
Low \citealt{mel13}).

Traditional radio correlators cross-multiply the signals from all pairs of antennas, and 
the computation scales as the number of antennas squared, $\mathcal{O}(\Nant^2)$ 
\citep{bun04}. As the number of elements in future arrays grows, the computational cost will 
become prohibitively expensive, and exploring efficient correlator schemes is essential to 
enable next generation instruments \citep{lon00}. Meanwhile, radio transient monitoring 
requires access to high time and frequency resolution data to identify and characterize events 
such as fast radio bursts (FRBs, \citealt{lor07}), or to follow up gravitational wave candidates 
with radio observations \citep{abb16a,abb16b}. FRBs are highly unexplored at low frequencies 
(< 1 GHz), but are expected to occur on timescales $\Delta t \sim$ 1--10~ms \citep{tho13}. 
Recording the full visibility matrix for $\Nant \gtrsim 10^3$ arrays at this timescale leads to 
extremely high data write rates.

Direct imaging correlators are a new variety of radio correlator which aim to alleviate both the 
computational strain of forming $\Nant^2$ correlations and the high data throughput associated 
with short timescale science. This is done by performing a spatial fast Fourier transform (FFT) 
to image the antenna signals, then squaring and averaging in time. This process scales as 
$\mathcal{O}(\Ng \log_2 \Ng)$, where \Ng~is the number of grid points in the FFT \citep{mor11,
 teg09, teg10}. For certain classes of telescopes, significantly those envisioned for next 
 generation cosmology experiments, this scaling is a large improvement over the $\Nant^2$ 
 scaling of traditional methods. Furthermore, because images are generated in real-time, the native 
 output bandwidth will be lowered in cases where $\Ng < \Nant^2$, and has the potential to be 
 lowered even further with real-time transient processing.

@@

Here we present the E-field Parallel Imaging Calibration (EPICal) algorithm -- a novel solution 
to the calibration problem, which can be integrated into direct imaging correlators and scales 
only as the number of antennas, $\mathcal{O}(\Nant)$. This method uses a correlation of the 
uncalibrated antenna signal stream with an output image pixel from the backend of the 
correlator to solve for the complex gains of the antennas. Because the calibration must be 
applied before gridding and imaging, our solution requires an iterative approach where the data 
from one time series is used to update the gains which are applied to the following time series. 
An example implementation of the algorithm is available with the EPIC software 
package\footnote{http://github.com/nithyanandan/EPIC}.

We review the MOFF algorithm and derive the calibration algorithm in \S \ref{sec:math}. We 
then demonstrate the algorithm in simulations in \S \ref{sec:sim}, 
and discuss noise trends in \S \ref{sec:noise}. We then apply the algorithm to a sample LWA 
data set in \S \ref{sec:data}.
Finally we conclude and discuss potential extensions to the 
algorithm in \S \ref{sec:discussion}.

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{fig1.pdf}
\caption{Data flow for the MOFF algorithm, reproduced from \citealt{thy15c}. The time ordered 
electric field measurements are collected by the antenna elements then frequency transformed in the 
F-engine. The calibration is applied, then the electric field measurements are gridded using the 
antenna aperture illumination patterns. The imager performs a spatial 2D FFT on the gridded data before 
squaring and averaging. The accumulated images are then written to disk.}
\label{fig:moff_flow}
\end{center}
\end{figure}

@@

Next we turn to the measurement performed by an antenna with far-field radiation pattern 
$\mathcal{W}_a(\hat{\mathbf{s}})$, where the subscript $a$ denotes a specific antenna. We 
adopt the common convention of normalizing the radiation pattern to have response of 1 at the 
center of the main lobe \citep[e.g.][]{nap99}. Consequently, the translation between the terminal 
voltages to electric field measurements has been absorbed in our definition of the radiation 
pattern. The ideal antenna electric field measurement can be expressed as 
\begin{equation}
E^I_a(f,t) = \int \mathcal{W}_a(\hat{\mathbf{s}})\mathcal{E}(\hat{\mathbf{s}},f,t) e^{-2\pi i \mathbf{r} \cdot \hat{\mathbf{s}}}\, d\Omega. 
\end{equation}
Equivalently we can express the effect of the antenna radiation pattern in the plane of the 
aperture.
\begin{equation}
E^I_a(f,t) = \int_{\mathrm{aperture}}W_a(\mathbf{r}-\mathbf{r}_a)E_a(\mathbf{r},f,t)\,d^2\mathbf{r}
\end{equation}
Here we used the Fourier relationship between the antenna's far-field radiation pattern and the 
aperture illumination pattern (sometimes referred to as the aperture distribution), $W_a(\mathbf{r}-\mathbf{r}_a)=\int  \mathcal{W}_a(\hat{\mathbf{s}}) \, d
\Omega$. 
While it is possible to define each antenna aperture illumination pattern in a way to absorb the 
relative position of the antenna, we keep the $\mathbf{r}_a$ coordinate to retain the separable
concepts of the primary beam (the product of two antenna radiation patterns) and the array
point spread function (square of the Fourier transform of the array layout).

@@

The next step is to grid the calibrated electric field measurements to a finite set of regularly spaced 
positions on the 2D plane of the ground, $\mathbf{r}_j$.
The extent of the grid is chosen to encompass the footprint of the antenna array.
MOFF uses the antenna illumination patterns as the gridding 
kernel according to the software holography/A-transpose technique \citep{mor09,bha08}. 
Mathematically we can express the gridding operation as
\begin{equation}\label{eq:gridding}
E'_j=\sum_a W_a(\mathbf{r}_j-\mathbf{r}_a) E'_a(t)
\end{equation}
Formally each pixel is calculated by summing over all antennas and their illumination patterns.
However the illumination patterns are typically compact and non-overlapping, so most pixels
will only contain contribution from at most one antenna, allowing equation~\ref{eq:gridding}
to be calculated through a sparse matrix multiplication.
This gridding 
achieves an optimal map \citep{teg97b} and places the data on a regular grid. Unlike 
traditional correlators which grid visibilities, the MOFF grids the electric field measurements directly. This 
operation must be performed for every time sample, before any averaging of the data.

@@

We conclude this section by connecting our calibration expression to that found in a visibility 
framework. In the limit of a single bright calibrating source at center, we can greatly 
simplify equations~\ref{eq:Cna} and~\ref{eq:cal_solution}. We will assume the radiation patterns are 
normalized such that $\mathcal{W}(0)=1$. We can further drop the exponential phase terms 
because $\spix$ is perpendicular to our planar antenna array. We then absorb the true gains into the true visibilities in 
equation~\ref{eq:Cna} to express as a sum of measured, uncalibrated, visibilities.
\begin{equation}
\mathcal{C}^{(n)}_{a,0} \rightarrow \frac{1}{\Nant}\sum_b \frac{1}{g^{*(n)}_b} V_{ab}
\end{equation}

We next plug this expression into equation~\ref{eq:cal_solution} to find our simplified 
calibration solution for a single bright point source. Because our sky is a single bright point 
source, the model visibilities are simply the flux of the source, $S_{\mathrm{src}}$.
\begin{equation}
g^{(n+1)}_a \rightarrow \frac{\sum_b  V_{ab}/g^{*(n)}_b}{\Nant S_{\mathrm{src}}}
\end{equation}
This is simply a gain-weighted sum of the measured visibilities over the flux of the source, 
which is indeed the limiting result from a visibility approach, for example seen in \citealt{mit08}. 
The ability to recover the equivalent expression despite not actually forming the visibilities is a 
result of the fact that only sums over visibilities come into the visibility-based solution, as was described in 
\citealt{mor11}. We have confirmed the limiting case equivalence here, and will explore the 
more general case in more detail in the following sections.

@@

The calibration loop continues by updating the gain estimates every \caliter~time samples. The 
phases of our gain estimates are shown in figure~\ref{fig:sim_phase} for 20 such iterations. 
The phase error plotted is the phase relative to the true gain for each antenna (various colored 
lines). One antenna was used as a reference to fix the absolute phase, so has zero phase 
error. The other 50 antennas are shown to have error spanning $2\pi$ initially, and after about 
10 iterations lock into a solution, settling down to noise levels around iteration 12 (0.12~s). We 
stop the simulation when the updated gains trace the thermal noise of the simulated sources, 
which can be seen by the coherence of the 50 antenna gains after iteration 12.

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{fig3.pdf}
\caption{Phase error of gain estimates as a function of iteration for simulated calibration. The 
gains were initialized with random phases, but the calibration loop was able to recover the 
correct phases after about 12 iterations. Each line represents an antenna in our 51 MWA 
antenna sample.
}
\label{fig:sim_phase}
\end{center}
\end{figure}

The estimated gain amplitudes for the simulations are shown in figure~\ref{fig:sim_amp}. The 
quantity plotted is the magnitude of the estimated gains over the true gains, $\left|g^{(n)}_a/
g_a\right|$, which places all antennas on the same scale. We can see the amplitudes 
converge toward their true values around the same time as the phases (iteration $\sim$12). At 
the beginning of calibration we can see the importance of the damping factor. At $n=0$, a couple of 
gains are shown to have abnormally high amplitude estimates, notably one about 3.3 times its 
true value (red line). These unbalanced high estimates caused the entire set of gains to be 
under estimated at $n=1$, even with a damping factor of 0.35. By $n=5$ the unbalanced 
amplitudes have been damped out and the calibration continues. Without the damping factor, 
the oscillation seen in the first couple iterations would have been significantly larger and taken 
much longer to fade out.

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{fig4.pdf}
\caption{Gain estimate amplitudes as a function of iteration for simulated calibration. Again, 
each line represents an antenna in the 51 MWA antenna sample. The gain estimates were 
initialized randomly, while the true values were unity. After about 12 iterations we see the 
calibration loop has settled around the correct values, with only noise remaining.}
\label{fig:sim_amp}
\end{center}
\end{figure}

After the gains have converged, we see both the phases and amplitudes to continue to 
fluctuate coherently. This is due to the stochastic fluctuations of the sources 
themselves, which are simulated as zero mean random processes with variance proportional
to their flux density.
For this simulation we restricted each calibration 
iteration to only 10~ms integration time. In principle a calibration implementation could use 
short integration times to allow the gains to converge, then increase the integration time to 
reduce this noise.

Images created at the beginning of calibration and at the end are shown in 
figure~\ref{fig:sim_images}. Each image is obtained over 400 time sample integration, corresponding to 
all snapshot images created with a given set of gain estimates. The left panel shows the image 
produced with our initialized unity gains. Because the phases are completely random, the 
image is essentially noise with the primary beam evident. After 20 calibration iterations, the image is far 
more clear, shown in the middle panel. 
The ten sources are visible, along with rumble throughout the image. By comparing with
a perfectly calibrated image (right panel), we see that the rumble and the shapes of the
point sources are matched between the two images. We therefore conclude that the 
artifacts visible are dominated by the point spread function, rather than calibration errors.

\begin{figure*}
\begin{center}
\includegraphics[width=.3\linewidth]{fig5a.pdf}
\includegraphics[width=.3\linewidth]{fig5b.pdf}
\includegraphics[width=.3\linewidth]{fig5c.pdf}
\caption{Images formed during simulated calibration. \emph{Left:} An image generated from a 
single frequency channel and 400 time sample integration after the gain estimates are randomized. As 
expected with random phases, the image is completely noisy, with the shape of the primary 
beam evident. \emph{Middle:} An image formed after calibration, again with a single frequency 
channel and 400 time sample integration. Now the simulated point sources are visible in the image. 
\emph{Right:} An image generated from the same simulated data as the middle image, but 
with perfect calibration correction. Many of the features seen in the middle image are also
seen here, confirming that they are point spread function artifacts. The ten simulated (and modeled) point sources are highlighted with red circles.
\label{fig:sim_images}
}
\end{center}
\end{figure*}

\section{Noise trends}\label{sec:noise}
Next we study the effect of noise in our system, and the consequences 
of an incomplete sky model. We run a suite of simulations varying the receiver noise and 
integration times, while forming antenna-pixel correlations, $\Cna$, for EPICal gain solutions, 
and simultaneously forming visibilities from the same electric field measurement streams to find visibility-based 
gain solutions for comparison.

@@

To simulate receiver noise, we add a gaussian distributed complex random number to each 
antenna electric field measurement at each time sample according to 
equation~\ref{eq:apply_gain}. The level of the receiver noise, $\sigma_r = \left<\left|
n_a(f,t)\right|^2\right>$, is varied in different simulation runs. We include two limiting 
cases, where the receiver noise power is subdominant to the sky power ($\sigma_r = 
10.0$~Jy), and where the receiver dominates the total power ($\sigma_r=100.0$~Jy).

@@

We simultaneously form simulated visibilities, $V_{ab}$, by correlating all pairs of antenna electric field 
measurements (including the receiver noise). The electric field measurements are correlated for a duration equal to 
the effective integration time of the EPICal loop for comparison. We then find the 
visibility-based gain estimates by minimizing
\begin{equation}\label{eq:vis_cal}
\chi^2 = \sum_a\sum_{b\ne a} \left|V_{ab}-g_a g_b^* V^M_{ab}\right|^2,
\end{equation}
using both versions of the model visibilities described above.

@@

We made several choices while studying the behavior of the LWA data to improve our 
calibration. Through our previous imaging work, we noted that the flux scale of uncalibrated 
images was consistent with average gain amplitudes of roughly 0.25. To allow the calibration to 
converge quickly we initialized our gain estimates at this level. 
We also found a boost in signal to noise is achieved easily by assuming the gains are
constant across a range of frequencies, and averaging solutions across channels.
Here we average solutions across 300 channels, or about 58.6 kHz. 
With a fractional bandwidth $B/f_0 = 7.9 \times 10^{-4} \ll 1$, we assume a smooth bandpass 
across the band. 
A damping factor of $\gamma = 0.7$ was adopted.
The large damping factor, relative to what was used in simulations, was necessary due to
the larger noise levels in the LWA data compared to the ideal simulations.

@@

Figure~\ref{fig:data_phase} shows the phase of our gain estimates over 30 calibration 
iterations, again with each colored line representing a different antenna. Given the quality of 
uncalibrated image demonstrated in \cite{thy15c}, it is a bit surprising to see the phase 
variation in our solutions. However, the phases are relatively flat after about 15 iterations 
(modulo noise), and exhibit a central ``trunk" where the majority of phases are congregated. 
This behavior is suggestive that while the uncalibrated data were able to produce a viable 
image, the minor changes from our solutions will focus the image and improve the quality. The 
actual location of the ``trunk" (slightly negative) is simply determined by the reference antenna 
chosen to have identically zero phase, but happens to be slightly more positive than the bulk of 
antennas. For plotting clarity, we unwrapped phases resulting in phases that appear to exceed 
$\pm \pi$.

@@

To estimate the fractional gain noise of our solutions we calculate the standard deviation of the gain
estimates after iteration 15 for each antenna separately. We then divide these standard
deviations by the respective mean gain amplitudes. This resulted in a distribution of fractional
gain noise centered at 0.15 and a width of 0.04 ($\approx 0.15/\sqrt{15}$). This noise level
is consistent with what we expected from the simulations.
There were four outliers with noise levels significantly outside the distribution 
($\sigma_g \approx 0.29$). Upon closer inspection the gain amplitudes of these outliers had
converged later in the calibration loop, after iteration 15.

@@

\textbf{Improving sky model.} In this work we used a perfect sky model (for simulation), or a 
very simple sky model (for LWA data). In principle the images produced by the EPIC correlator 
can be used to improve the sky model used in calibration -- similar to a major loop in 
self-calibration. This can be especially useful for compact, widefield arrays which require a model of 
both compact and diffuse sources over a large patch of sky. In addition to requiring the sky 
model, the calibration also requires this model to be attenuated by the primary beam, which 
can be difficult to measure at the precision necessary \citep[e.g.][]{neb15,vir14,thy15b}. 
However, the direct imaging correlator provides exactly the model necessary in real time, and 
can be iterated over to improve the images and gain solutions.
